Graph Topological Aspects of Granger Causal Network Learning

### 2019-07-17 paper_theory/paper_theory.tex
I can have a supplementary document with proofs and extra figures.

\documentclass{statsoc}
% \usepackage{mathtime}
\usepackage{ccfonts,eulervm}
% \usepackage[T1]{fontenc}
% \usepackage[lite]{belleek}
% \fontfamily{belleek}
% \usepackage[math]{belleek}
% \usepackage{concretemath-fonts}

% ---------------------------------------------------------------------------------------------------------------------------------
% The statsoc style file is broken?
% I have copied this code straight from stack overflow
% https://tex.stackexchange.com/questions/351768/problem-with-statsoc-class-and-pdflatex
\usepackage{geometry}

@@

%% bibliography stuff -- this needs come before the preamble inclusion
\usepackage{enumitem}
\usepackage{etex,etoolbox}
\usepackage{hyperref}
\usepackage{fullpage}
\setlength{\marginparwidth}{2cm}
\usepackage{todonotes}

@@

% Math packages, theorem definitions and numbering
\usepackage{amsmath}
\usepackage{amssymb}
% \usepackage{amsthm}
% \usepackage{mathrsfs} % Fancy scripted font
% \usepackage{bm}  % Bold math

@@

\clearpage

% \tableofcontents
\clearpage

@@

\subsection{Contributions}
In the context of time series data, sparsity assumptions remain
important, but there is significant additional structure that may
arise as a result of considering the topology of the underlying
Granger-causality graph, which to our knowledge remains largely
unexplored.  The focus of this paper is to shed light on some of these
topological questions, in particular, we study a particularly simple
notion of causality graph topology which we term ``strongly causal''
and show that stationary times series whose underlying causality graph
has this structure satisfy natural intuitive notions of ``information
flow'' through the graph.  Moreover, we show that such graphs are
perfectly recoverable with only \textit{pairwise} Granger-causality
tests, which would otherwise suffer from serious confounding problems.
Aside from being an interesting theoretical perspective, prior
assumptions about the underlying graph (similarly to sparsity
assumptions) can greatly improve upon the statistical power of
causality graph recovery algorithms when the assumptions are met.

@@

The principle contributions of this paper are as follows: firstly, in
section \ref{sec:theory} we study \textit{pairwise} Granger-causality
relations, providing novel theorems connecting the structure of the
causality graph to the pairwise ``causality flow'' in the system, as
well as an interpretation in terms of the graph topology of the
sparsity pattern of matrices arising in the Wold decomposition,
generalizing in some sense the notion of ``feedback-free'' processes
studied by \cite{caines1975feedback} in close connection with
Granger-causality.  We establish sufficient conditions (sections
\ref{sec:strongly_causal_graphs}, \ref{sec:persistent_systems}) under
which a fully conditional Granger-causality graph can be recovered
from pairwise tests alone (sec \ref{sec:pairwise_algorithm}).  We
report \hl{elsewhere}\todo{where?  Or can I fit in a brief summary?}
the results of extensive finite sample simulations as well as an
application to EEG data, where in addition we summarize appropriate
methods for hypothesis testing, model order selection, computationally
efficient estimations, and error rate controls in practice.  Our
simulation results establish significant potential for improvement
over existing methods.  Concluding remarks on further open problems
and extentions are provided in Section \ref{sec:conclusion}.

@@

As a consequence of the Wold decomposition theorem \cite{lindquist},
every WSS sequence has the moving average $MA(\infty)$
representation

@@

The notion of conditional orthogonality is used throughout, as well as
some equivalent representations for Granger-causality.

\begin{lemma}[\cite{lindquist} Proposition 2.4.2]
  \label{lem:conditional_orthogonality_equivalence}
  Consider three closed subspaces of a Hilbert space $\mathcal{A}$,
  $\mathcal{B}$, $\mathcal{X}$.  The following statements are
  equivalent

  \todo{Do I even use (a) anymore?  I think I can just put the definition of conditional orthogonality into Theorem 1.}

  \begin{enumerate}
    \item{$\mathcal{A} \perp \mathcal{B}\ |\ \mathcal{X}$}
    \item{$\linE{\beta}{\mathcal{A} \vee \mathcal{X}} = \linE{\beta}{\mathcal{X}}\ \forall \beta \in \mathcal{B}$.}
    % \item{$\linE{\beta}{\mathcal{A}} = \linE{\linE{\beta}{\mathcal{X}}}{\mathcal{A}}\ \forall \beta \in \mathcal{B}}$}
    \end{enumerate}

    Where $\mathcal{A} \perp \mathcal{B}\ |\ \mathcal{X}$ denotes
    conditional orthogonality:

    \begin{equation*}
      \inner{a - \linE{a}{\mathcal{X}}}{b - \linE{b}{\mathcal{X}}} = 0\ \forall a \in \mathcal{A}, b \in \mathcal{B}.
    \end{equation*}
\end{lemma}

@@

\begin{proof}
  % Geweke uses the log of the ratio of the determinants of the residual variances
  % 
  % The equivalence $(1) \iff (3)$ is essentially a restatement of the
  % definition, but is in line with the seminal work of Geweke
  % \cite{geweke1982measurement}, \cite{geweke1984}.

  $(1) \Rightarrow (2)$ follows as a result of the uniqueness of orthogonal
  projection (i.e. the best estimate is necessarily the coefficients
  of the model).  $(2) \Rightarrow (3)$ follows since in computing
  $(y - \linE{y}{\H_{t - 1}^{-j}})$ for $y \in H_t^i$ it is sufficient
  to consider $y = x_i(t)$ by linearity, then since
  $H_{t - 1}^i \subseteq \H_{t - 1}^{-j}$ we have
  $(x_i(t) - \linE{x_i(t)}{\H_{t = 1}^{-j}}) = v_i(t)$ since
  $\B_{ij}(z) = 0$ and $v_i(t) \perp \H_{t - 1}$.  $(3) \iff (4)$ is a result
  of Lemma \ref{lem:conditional_orthogonality_equivalence}.  And,
  $(4) \implies (1)$ follows directly from the Definition.
\end{proof}

@@

The edges of the Granger-causality graph $\gcg$ can be given a general
notion of ``weight'' by associating an edge $(j, i)$ with the
\textit{strictly causal} LTI filter $\B_{ij}(z)$ (see eqn
\eqref{eqn:filter_action}).  Thence, the matrix $\B(z)$ is analogous
to a \textit{weighted adjacency matrix}\footnote{We are using the
  convention that $\B_{ij}(z)$ is a filter with input $x_j$ and output
  $x_i$ so as to write the action of the system as $\B(z)x(t)$ with
  $x(t)$ as a column vector.  This competes with the usual convention
  for adjacency matrices where $A_{ij} = 1$ if there is an edge
  $(i, j)$.  In our case, the sparsity pattern of $\B_{ij}$ is the
  \textit{transposed} conventional adjacency matrix.} for the graph $\gcg$.  And,
in the same way that the $k^{\text{th}}$ power of an adjacency matrix
counts the number of paths of length $k$ between nodes,
$(\B(z)^k)_{ij}$ is a filter isolating the ``action'' of $j$ on $i$ at
a time lag of $k$ steps, this is exemplified in the inversion formula
\ref{eqn:lsi_inversion}.

An elementary theorem connecting the adjacency matrix with paths will
allow us to deduce the sparsity pattern of $\A(z)$.  Proof follows
easily by induction:

\begin{lemma}
  \label{lem:adj_matrix}
  Let $S$ be the transposed adjacency matrix of the Granger-causality
  graph $\gcg$.  Then, $(S^k)_{ij}$ is the number of paths of length
  $k$ from node $j$ to node $i$.  Evidently, if
  $\forall k \in \N,\ (S^k)_{ij} = 0$ then $j \not\in \anc{i}$.
\end{lemma}

@@

\begin{proof}
  Equation \eqref{eqn:parent_expansion} is immediate from the
  $\VAR(\infty)$ representation of \eqref{eqn:ar_representation} and
  Theorem \ref{thm:granger_causality_equivalences}, we are left to
  demonstrate \eqref{eqn:ancestor_expansion}.

  From equation (\ref{eqn:ar_representation}), which we are assuming
  throughout the paper to be invertible, we can write

  \begin{equation*}
    x(t) = (I - \B(z))^{-1} v(t),
  \end{equation*}

  where $(I - \B(z))^{-1} = \A(z)$ due to the uniqueness of
  (\ref{eqn:wold}).  Since $\B(z)$ is stable we have

  \begin{equation}
    \label{eqn:resolvant_inv}
    (I - \B(z))^{-1} = \sum_{k = 0}^\infty \B(z)^k.
  \end{equation}

  Invoking the Cayley-Hamilton theorem allows writing the infinite sum
  of \eqref{eqn:resolvant_inv} in terms of \textit{finite} powers of
  $\B$.

  Let $S$ be a matrix with elements in $\{0, 1\}$ which represents the
  sparsity pattern of $\B(z)$, from lemma \ref{lem:adj_matrix} $S$ is
  the transpose of the adjacency matrix for $\gcg$ and hence
  $(S^k)_{ij}$ is non-zero if and only if $j \in \gpn{k}{i}$, and
  therefore $\B(z)^k_{ij} = 0$ if $j \not \in \gpn{k}{i}$.  Finally,
  since $\anc{i} = \bigcup_{k = 1}^n\gpn{k}{i}$ we see that
  $\A_{ij}(z)$ is zero if $j \not\in \anc{i}$.

  Thence

  \begin{align*}
    x_i(t) &= [(I - \B(z))^{-1}v(t)]_i = \sum_{j = 1}^n \A_{ij}(z) v_j(t)\\
    &= \A_{ii}(z) v_i(t) + \sum_{\substack{j \in \anc{i} \\ j \ne i}} \A_{ij}(z) v_j(t)
  \end{align*}
\end{proof}

@@

\begin{lemma}
  \label{lem:separated_ancestor_uncorrelated}
  Consider distinct nodes $i, j$ in a Granger-causality graph
  $\gcg$.  If

@@

  then $\H_t^{(i)} \perp \H_t^{(j)}$, that is,
  $\forall s, \tau \in \Z_+\ \E[x_i(t - s)x_j(t - \tau)] = 0$.  Moreover,
  this means that $j \npwgc i$ and $\linE{x_j(t)}{\H_{t - 1}^i} = 0$.
\end{lemma}

@@

\begin{proof}
  We show directly that
  $\forall s, \tau \in \Z_+\ \E[x_i(t - s)x_j(t - \tau)] = 0$.  To this end,
  fix $s, \tau \ge 0$, then by expanding with equation \eqref{eqn:ancestor_expansion} we have

  \begin{align*}
    \E x_i(t - s)x_j(t - \tau)
    &= \E \big(\A_{ii}(z)v_i(t - s)\big)\big(\A_{jj}(z)v_j(t - \tau)\big)\\
    &+ \sum_{\substack{k \in \anc{i} \\ k \ne i}}\E[\big(\A_{ik}(z)v_k(t - s)\big)\big(\A_{jj}(z)v_j(t - \tau)\big)]\\
    &+ \sum_{\substack{k \in \anc{j} \\ k \ne j}}\E[\big(\A_{ii}(z)v_i(t - s)\big) \big(\A_{jk}(z) v_k(t - \tau)\big)]\\
    &+ \sum_{\substack{k \in \anc{i} \\ k \ne i}}\sum_{\substack{\ell \in \anc{j} \\ \ell \ne j}}\E[\big(\A_{ik}(z)v_k(t - s)\big)\big(\A_{j\ell}(z)v_\ell(t - \tau)\big)].
  \end{align*}

  Keeping in mind that $v(t)$ is an isotropic and uncorrelated
  sequence we see that each of these above four terms are 0: the
  first term since $i \ne j$, the second and third since
  $j \not\in \anc{i}$ and $i \not\in \anc{j}$ and finally the fourth since
  $\anc{i} \cap \anc{j} = \emptyset$.
\end{proof}

It is possible for components of $x(t)$ to be correlated at some time
lags without resulting in pairwise causality:

\begin{lemma}
  \label{lem:ancestor_uncorrelated}
  Consider distinct nodes $i, j$ in a Granger-causality graph $\gcg$.
  If

@@

  then $j \npwgc i$.
\end{lemma}
\begin{proof}
  By Theorem \ref{thm:granger_causality_equivalences} it suffices to show that

  \begin{equation*}
    \inner{x_i(t) - \linE{x_i(t)}{\H_{t - 1}^i}}{x_j(t) - \linE{x_j(t)}{\H_{t - 1}^i}} = 0,
  \end{equation*}

  which by the orthogonality principle is equivalent to

  \begin{equation}
    \label{eqn:proof_inner0}
    \inner{x_i(t)}{x_j(t) - \linE{x_j(t)}{\H_{t - 1}^i}} = 0.
  \end{equation}

  Define the disjoint sets

  \begin{align*}
    C_0(u) &= \{k \in \pa{u}\ |\ i \not\in \anc{k}, k \ne i \}\\
    C_1(u) &= \{k \in \pa{u}\ |\ i \in \anc{k} \text{ or } k = i \}.
  \end{align*}

  We can then expand the parents of $j$ using Equation \ref{eqn:parent_expansion} as

  \begin{equation*}
    x_j(t) = v_j(t) + \sum_{k \in C_0(j)} B_{jk}(z) x_k(t) + \sum_{k \in C_1(j)} B_{jk}(z) x_k(t),
  \end{equation*}

  which when substituted into the left hand side of Equation
  \ref{eqn:proof_inner0} results (by Lemma
  \ref{lem:separated_ancestor_uncorrelated}) in

  \begin{equation*}
    \inner{x_i(t)}{\sum_{k \in C_1(j)} B_{jk}(z) x_k(t) - \sum_{k \in C_1(j)} \linE{B_{jk}(z) x_k(t)}{\H_{t - 1}^i}}.
  \end{equation*}

  We can continue this process recursively (i.e. split each
  $k \in C_1(j)$ into $C_0(k)$ and $C_1(k)$) which must eventually
  terminate with $C_1(u) = \{i\}$.  Therefore, there exists some
  causal filter $\Phi(z)$ such that Equation \ref{eqn:proof_inner0} is
  equivalent to

  \begin{equation}
    \label{eqn:proof_inner0}
    \inner{x_i(t)}{\Phi(z)x_i(t) - \linE{\Phi(z)x_i(t)}{\H_{t - 1}^i}},
  \end{equation}

  which is $0$ since $\linE{\Phi(z)x_i(t)}{\H_{t - 1}^i} = \Phi(z)x_i(t)$.
\end{proof}

The previous Lemma can be strengthened significantly; notice
that it is possible to have some $k \in \anc{i} \cap \anc{j}$ where still
$j \npwgc i$, an example is furnished by the three node graph
$k \rightarrow i \rightarrow j$ where clearly
$k \in \anc{i}\cap\anc{j}$ but $j \npwgc i$.  We must introduce the concept
of a \textit{confounding} variable, which effectively eliminates the
possibility presented in this example.

@@

\begin{proof}
  We will prove by way of contradiction.  To this end, suppose that
  $j$ is a node such that: $(a)$ $j \not \in \anc{i}$ and $(b)$ for
  every $k \in \anc{i} \cap \anc{j}$ every
  $k \rightarrow \cdots \rightarrow j$ path contains $i$.

  Firstly, notice that every $u \in \big(\pa{j} \setminus \{i\}\big)$
  necessarily inherits these same two properties.  This follows since
  if we also had $u \in \anc{i}$ then $u \in \anc{i} \cap \anc{j}$ so that every
  $u \rightarrow \cdots \rightarrow j$ path must contain $i$, but
  $u \in \pa{j}$, so this is not the case since $u \rightarrow j$ is a
  path that doesn't contain $i$; moreover, if we consider
  $w \in \anc{i} \cap \anc{u}$ then we also have
  $w \in \anc{i} \cap \anc{j}$ so every
  $w \rightarrow \cdots \rightarrow j$ path must contain $i$.  These
  properties therefore extend inductively to every
  $u \in \big(\anc{j} \setminus \{i\}\big)$.

  In order to deploy a recursive argument, define the following
  partition of $\pa{u}$, for some node $u$:

  \begin{align*}
    C_0(u) &= \{k \in \pa{u}\ |\ i \not\in \anc{k}, \anc{i} \cap \anc{k} = \emptyset, k \ne i\}\\
    C_1(u) &= \{k \in \pa{u}\ |\ i \in \anc{k} \text{ or } k = i\}\\
    C_2(u) &= \{k \in \pa{u}\ |\ i \not\in \anc{k}, \anc{i} \cap \anc{k} \ne \emptyset, k \ne i\}.
  \end{align*}

  We notice that for any $u$ having the properties $(a), (b)$ above, we
  must have $C_2(u) = \emptyset$ since if $k \in C_2(u)$ then
  $\exists w \in \anc{i} \cap \anc{k}$ s.t. $i \not \in \anc{k}$ and
  therefore there must be a path $\gcgpath{w}{k} \rightarrow u$ which
  does not contain $i$.

  Using this partition, we will expand $x_j(t)$ in terms of it's
  parents, and recursively expand nodes in $C_1$ until we reach a case
  where $C_1 = \emptyset$.  For the first step equation
  \eqref{eqn:parent_expansion} gives us:

  \begin{equation}
    \label{eqn:xj_partition_expansion}
    x_j(t) = \A_{jj}(z)\Big(v_j(t) + \sum_{k \in C_0(j)}\B_{ik}(z)x_k(t) + \sum_{k \in C_1(j)}\B_{ik}(z)x_k(t)\Big).
  \end{equation}

  Using this representation we choose an arbitrary $\Phi(z) x_i(t - 1) \in \H_{t - 1}^{(j)}$ and show that

  \begin{equation}
    \inner{x_i(t) - \linE{x_i(t)}{\H_{t - 1}^{(i)}}}{\Phi(z)x_j(t - 1) - \linE{\Phi(z)x_j(t - 1)}{\H_{t - 1}^{(i)}}} = 0,
  \end{equation}

  which will imply (by Theorem \ref{thm:granger_causality_equivalences}) that $j \npwgc i$ and for which it is equivalent to show that

  \begin{equation}
    \label{eqn:sufficient_inner_prod}
    \inner{x_i(t)}{\Phi(z)x_j(t - 1) - \linE{\Phi(z)x_j(t - 1)}{\H_{t - 1}^{(i)}}} = 0,
  \end{equation}

  by the orthogonality principle since $\linE{x_i(t)}{\H_{t - 1}^{(i)}} \in \H_{t - 1}^{(i)}$.  Substituting (\ref{eqn:xj_partition_expansion}) into (\ref{eqn:sufficient_inner_prod}) and starting with the first term we have

  \begin{align*}
    &\inner{x_i(t)}{\Phi(z)\A_{jj}(z)v_j(t - 1) - \linE{\Phi(z)\A_{jj}(z)v_j(t - 1)}{\H_{t - 1}^{(i)}}}\\
    \overset{(\alpha)}{=}\ &\inner{\A_{ii}(z)v_i(t) + \sum_{k \in \anc{i}}\A_{ik}(z)v_k(t)}{\Phi(z)\A_{jj}(z)v_j(t - 1)}\\
    \overset{(\beta)}{=}\ &0,
  \end{align*}

  where $(\alpha)$ follows by expanding $x_i(t)$ with (\ref{eqn:ancestor_expansion}) and $\linE{\Phi(z)\A_{jj}(z)v_j(t - 1)}{\H_{t - 1}^{(i)}} = 0$ because $\forall \tau, s$

  \begin{equation*}
    \E v_j(t - \tau) x_i(t - s) = \E v_j(t - \tau) \sum_{k \in \anc{i} \cup \{i\}}\A_{ik}(z)v_k(t - s) = 0,
  \end{equation*}

  since $j \not \in \anc{i}$; $(\beta)$ follows similarly, that is, $j \not \in \anc{i}$.  Secondly we see that $\forall k \in C_0(j)$

  \begin{equation*}
    \inner{x_i(t)}{\Phi(z)\A_{jj}(z)\B_{ik}(z)x_k(t - 1) - \linE{\Phi(z)\A_{jj}(z)\B_{ik}(z)x_k(t - 1)}{\H_{t - 1}^{(i)}}} = 0,
  \end{equation*}

  which follows from Lemma \ref{lem:ancestor_uncorrelated}.  Finally,
  for $k \in C_1(j)$ the case $k = i$ is immediate (since the error in
  estimating $\Phi(z)\B_{ii}(z)x_i(t)$ given $\H_{t - 1}^{(i)}$ is
  $0$), so suppose $k \ne i$.  We know from above that $k$ inherits the
  key properties referred to as $(a)$ and $(b)$ above and therefore we
  can recursively expand $k$ in the same way as in equation
  (\ref{eqn:xj_partition_expansion}).  Continuing this recursion for
  each $k \in C_1(j)$ (where $k \ne i$) must eventually terminate since
  $i \in \anc{k}$.
\end{proof}

@@

% \begin{example}
%   \label{ex:diamond_cancellation}
%   Firstly, on $n = 4$ nodes, ``diamond'' shapes can lead to cancellation on paths of length 2:

% \begin{equation*}
%   x(t) =
%   \left[
%     \begin{array}{cccc}
%       0 & 0 & 0 & 0\\
%       a & 0 & 0 & 0\\
%       -a & 0 & 0 & 0\\
%       0 & 1 & 1 & 0\\
%     \end{array}
%   \right] x(t - 1) + v(t),
% \end{equation*}

% with $\E v(t) = 0,\ \E v(t)v(t - \tau)^\T = \delta_\tau I$.

% By directly calculating

% \begin{align*}
%   x_4(t) &= x_2(t - 1) + x_3(t - 1) + v_4(t)\\
%          &= ax_1(t - 2) + av_2(t - 1) - ax_1(t - 2) -av_3(t - 1) + v_4(t)\\
%          &= a(v_2(t - 1) - v_3(t - 1)) + v_4(t),
% \end{align*}

% we see that, since $v(t)$ is isotropic white noise, $1 \npwgc 4$.  The problem here is that there are multiple paths from $x_1$ to $x_4$.
% \end{example}

\begin{figure}
  \centering
  \begin{subfigure}[b]{0.35\textwidth}
    \includegraphics[width=\linewidth]{example1.pdf}
    \caption{Example multiple path cancellation: $j \in \anc{i} \nRightarrow j \pwgc i$}
    \label{fig:diamond_cancellation}
  \end{subfigure}
  \begin{subfigure}[b]{0.35\textwidth}
    \includegraphics[width=\linewidth]{example2.pdf}
    \caption{Example cancellation from time lag: $j \in \pa{i} \nRightarrow j \pwgc i$}
    \label{fig:lag_cancellation}
  \end{subfigure}
\end{figure}

% \begin{example}
%   \label{ex:lag_cancellation}
%   A second example on $n = 3$ nodes is also worth examining, in this case
%   cancellation is a result of differing time lags.

% \begin{equation*}
%   x(t) =
%   \left[
%     \begin{array}{ccc}
%       0 & 0 & 0\\
%       -a & 0 & 0\\
%       0 & 1 & 0\\
%     \end{array}
%   \right] x(t - 1) +
%   \left[
%     \begin{array}{ccc}
%       0 & 0 & 0\\
%       0 & 0 & 0\\
%       a & 0 & 0\\
%     \end{array}
%   \right] x(t - 2) + v(t)
% \end{equation*}

% Then

% \begin{align*}
%   x_2(t) &= v_2(t) - ax_1(t - 1)\\
%   x_3(t) &= v_3(t) + x_2(t - 1) + ax_1(t - 2)\\
%   \Rightarrow x_3(t) &= v_2(t - 1) + v_3(t),
% \end{align*}

% and again $1 \npwgc 3$.
% \end{example}

@@

\begin{definition}[Strongly Causal]
  \label{def:strongly_causal}
  We will say that a Granger-causality graph $\gcg$ is
  \textit{strongly causal} if there is at most 1 directed path between
  any two nodes.  Strongly Causal Graphs will be referred to as SCGs.
\end{definition}

@@

\begin{proposition}
  \label{prop:pwgc_anc}
  If $\gcg$ is a strongly causal DAG then $j \in \anc{i} \Rightarrow j \pwgc i$.
\end{proposition}
\begin{proof}
  We will show that for some $\psi \in \H_{t - 1}^{(j)}$ we have

>>>>

\documentclass{statsoc}
% \usepackage{mathtime}
\usepackage{ccfonts,eulervm}
\usepackage{tabularx}
\usepackage{xr}
\externaldocument{proofs}
% \usepackage[T1]{fontenc}
% \usepackage[lite]{belleek}
% \fontfamily{belleek}
% \usepackage[math]{belleek}
% \usepackage{concretemath-fonts}
% \interfootnotelinepenalty=10000
% ---------------------------------------------------------------------------------------------------------------------------------
% The statsoc style file is broken?
% I have copied this code straight from stack overflow
% https://tex.stackexchange.com/questions/351768/problem-with-statsoc-class-and-pdflatex
\usepackage{geometry}

@@

%% bibliography stuff -- this needs come before the preamble inclusion
\usepackage{enumitem}
\usepackage{etex,etoolbox}
\usepackage{hyperref}
\usepackage{fullpage}
\setlength{\marginparwidth}{2cm}
\usepackage{todonotes}
\usepackage{multirow}
\usepackage{booktabs}

@@

% Math packages, theorem definitions and numbering
\usepackage{amsmath}
\usepackage{amssymb}
% \usepackage{amsthm}
% \usepackage{mathrsfs} % Fancy scripted font
\usepackage{bm}  % Bold math

@@

% \clearpage

% % \tableofcontents
% \clearpage

@@

% \subsection{Contributions}
In the context of time series data, sparsity assumptions remain
important, but there is significant additional structure that may
arise as a result of considering the topology of the underlying
Granger-causality graph, which to our knowledge remains largely
unexplored.  The focus of this paper is to shed light on some of these
topological questions, in particular, we study a particularly simple
notion of causality graph topology which we term ``strongly causal''
and show that stationary times series whose underlying causality graph
has this structure satisfy natural intuitive notions of ``information
flow'' through the graph.  Moreover, we show that such graphs are
perfectly recoverable with only \textit{pairwise} Granger-causality
tests, which would otherwise suffer from serious confounding problems.
Aside from being an interesting theoretical perspective, prior
assumptions about the underlying graph (similarly to sparsity
assumptions) can greatly improve upon the statistical power of
causality graph recovery algorithms when the assumptions are met.

@@

The principle contributions of this paper are as follows: firstly, in
section \ref{sec:theory} we study \textit{pairwise} Granger-causality
relations, providing novel theorems connecting the structure of the
causality graph to the pairwise ``causality flow'' in the system, as
well as an interpretation in terms of the graph topology of the
sparsity pattern of matrices arising in the Wold decomposition,
generalizing in some sense the notion of ``feedback-free'' processes
studied by \cite{caines1975feedback} in close connection with
Granger-causality.  We establish sufficient conditions (sections
\ref{sec:strongly_causal_graphs}, \ref{sec:persistent_systems}) under
which a fully conditional Granger-causality graph can be recovered
from pairwise tests alone (sec \ref{sec:pairwise_algorithm}).  We
report a summary of simulation results in \ref{sec:simulation}, with
additional results reported in the supplementary material
\ref{apx:simulation}.  Our simulation results establish that there is
significant potential for improvement over existing methods, and that
the graph-topological aspects of time series analysis are relevant for
both theory and practice.  Concluding remarks on further open problems
and extentions are provided in Section \ref{sec:conclusion}.  The
proofs of each proposition and theorem are also relegated to the
supplementary material, simple corollaries have proofs included in the
main text.

@@

As a consequence of the Wold decomposition theorem (see e.g. \cite{lindquist}),
every WSS sequence has the moving average $MA(\infty)$
representation

@@

The notion of conditional orthogonality is the essence of
Granger-causality, and enables us to obtain results for a fairly
general class of WSS processes, rather than simply $\VAR(p)$ models.

\begin{definition}[Conditional Orthogonality]
  \label{lem:conditional_orthogonality_equivalence}
  Consider three closed subspaces of a Hilbert space $\mathcal{A}$,
  $\mathcal{B}$, $\mathcal{X}$.  We say that $\mathcal{A}$ is
  conditionally orthogonal to $\mathcal{B}$ given $\mathcal{X}$
  and write $\mathcal{A} \perp \mathcal{B}\ |\ \mathcal{X}$ if

    \begin{equation*}
      \inner{a - \linE{a}{\mathcal{X}}}{b - \linE{b}{\mathcal{X}}} = 0\ \forall a \in \mathcal{A}, b \in \mathcal{B}.
    \end{equation*}

  An equivalent condition is that (see \cite{lindquist} Proposition 2.4.2)
  \begin{equation*}
    \linE{\beta}{\mathcal{A} \vee \mathcal{X}} = \linE{\beta}{\mathcal{X}}\ \forall \beta \in \mathcal{B}
  \end{equation*}
\end{definition}

@@

% \begin{proof}
%   % Geweke uses the log of the ratio of the determinants of the residual variances
%   % 
%   % The equivalence $(1) \iff (3)$ is essentially a restatement of the
%   % definition, but is in line with the seminal work of Geweke
%   % \cite{geweke1982measurement}, \cite{geweke1984}.

%   $(a) \Rightarrow (b)$ follows as a result of the uniqueness of orthogonal
%   projection (i.e. the best estimate is necessarily the coefficients
%   of the model).  $(b) \Rightarrow (c)$ follows since in computing
%   $(y - \linE{y}{\H_{t - 1}^{-j}})$ for $y \in H_t^i$ it is sufficient
%   to consider $y = x_i(t)$ by linearity, then since
%   $H_{t - 1}^i \subseteq \H_{t - 1}^{-j}$ we have
%   $(x_i(t) - \linE{x_i(t)}{\H_{t = 1}^{-j}}) = v_i(t)$ since
%   $\B_{ij}(z) = 0$ and $v_i(t) \perp \H_{t - 1}$.  $(c) \iff (d)$ is a result
%   of the equivalence in Definition
%   \ref{lem:conditional_orthogonality_equivalence}.  And,
%   $(d) \implies (a)$ follows directly from the Definition.
% \end{proof}

@@

The edges of the Granger-causality graph $\gcg$ can be given a general
notion of ``weight'' by associating an edge $(j, i)$ with the
\textit{strictly causal} LTI filter $\B_{ij}(z)$ (see eqn
\eqref{eqn:filter_action}).  Thence, the matrix $\B(z)$ is analogous
to a \textit{weighted adjacency matrix}\footnote{\footnotesize We are using the
  convention that $\B_{ij}(z)$ is a filter with input $x_j$ and output
  $x_i$ so as to write the action of the system as $\B(z)x(t)$ with
  $x(t)$ as a column vector.  This competes with the usual convention
  for adjacency matrices where $A_{ij} = 1$ if there is an edge
  $(i, j)$.  In our case, the sparsity pattern of $\B_{ij}$ is the
  \textit{transposed} conventional adjacency matrix.} for the graph $\gcg$.  And,
in the same way that the $k^{\text{th}}$ power of an adjacency matrix
counts the number of paths of length $k$ between nodes,
$(\B(z)^k)_{ij}$ is a filter isolating the ``action'' of $j$ on $i$ at
a time lag of $k$ steps, this is exemplified in the inversion formula
\ref{eqn:lsi_inversion}.

@@

@@

The following propositions are essentially lemmas used for the proof
of the upcomming Proposition \ref{prop:ancestor_properties}, but
remain relevant for providing intuitive insight into the problems at
hand.

\begin{proposition}
  Consider distinct nodes $i, j$ in a Granger-causality graph
  $\gcg$.  If

@@

  then $\H_t^{(i)} \perp \H_t^{(j)}$, that is,
  $\forall s, \tau \in \Z_+\ \E[x_i(t - s)x_j(t - \tau)] = 0$.  Moreover,
  this means that $j \npwgc i$ and $\linE{x_j(t)}{\H_{t - 1}^i} = 0$.
\end{proposition}

@@

\begin{remark}
  It is possible for components of $x(t)$ to be correlated at some
  time lags without resulting in pairwise causality.  For instance,
  the conclusion $j \npwgc i$ of Proposition
  \ref{prop:separated_ancestor_uncorrelated} will still hold even if
  $i \in \anc{j}$, since $j$ cannot provide any information about $i$
  that is not available from observing $i$ itself.  However, it is
  important to note that this is the case only if we have direct
  access to $x(t)$.  That is, if we instead observed a noisy version
  of $x$ (i.e. $\tilde{x}(t) = x(t) + n(t)$) the observation of
  $\tilde{x}_j$ would enable conditionally improved estimates of
  $\tilde{x}_i$, without there necessarily being any causal flow in
  the underlying system $x(t)$.
\end{remark}

\begin{proposition}
  Consider distinct nodes $i, j$ in a Granger-causality graph $\gcg$.
  If

@@

  then $j \npwgc i$.
\end{proposition}

The previous result can still be strengthened significantly; notice
that it is possible to have some $k \in \anc{i} \cap \anc{j}$ where still
$j \npwgc i$, an example is furnished by the three node graph
$k \rightarrow i \rightarrow j$ where clearly
$k \in \anc{i}\cap\anc{j}$ but $j \npwgc i$.  We must introduce the concept
of a \textit{confounding} variable, which effectively eliminates the
possibility presented in this example.

@@

@@

\begin{example}
  \label{ex:diamond_cancellation}
  Firstly, on $n = 4$ nodes, ``diamond'' shapes can lead to cancellation on paths of length 2:

\begin{equation*}
  x(t) =
  \left[
    \begin{array}{cccc}
      0 & 0 & 0 & 0\\
      a & 0 & 0 & 0\\
      -a & 0 & 0 & 0\\
      0 & 1 & 1 & 0\\
    \end{array}
  \right] x(t - 1) + v(t),
\end{equation*}

with $\E v(t) = 0,\ \E v(t)v(t - \tau)^\T = \delta_\tau I$.

By directly calculating

\begin{align*}
  x_4(t) &= x_2(t - 1) + x_3(t - 1) + v_4(t)\\
         &= ax_1(t - 2) + av_2(t - 1) - ax_1(t - 2) -av_3(t - 1) + v_4(t)\\
         &= a(v_2(t - 1) - v_3(t - 1)) + v_4(t),
\end{align*}

we see that, since $v(t)$ is isotropic white noise, $1 \npwgc 4$.  The problem here is that there are multiple paths from $x_1$ to $x_4$.
\end{example}

\begin{figure}
  \centering
  \caption{Examples illustrating the difficulty of obtaining a converse to Proposition \ref{prop:ancestor_properties}}
  \begin{subfigure}[b]{0.35\textwidth}
    \includegraphics[width=\linewidth]{example1.pdf}
    \caption{Path cancellation: $j \in \anc{i} \nRightarrow j \pwgc i$}
    \label{fig:diamond_cancellation}
  \end{subfigure}
  \begin{subfigure}[b]{0.35\textwidth}
    \includegraphics[width=\linewidth]{example2.pdf}
    \caption{Cancellation from time lag: $j \in \pa{i} \nRightarrow j \pwgc i$}
    \label{fig:lag_cancellation}
  \end{subfigure}
\end{figure}

\begin{example}
  \label{ex:lag_cancellation}
  A second example on $n = 3$ nodes is also worth examining, in this case
  cancellation is a result of differing time lags.

\begin{equation*}
  x(t) =
  \left[
    \begin{array}{ccc}
      0 & 0 & 0\\
      -a & 0 & 0\\
      0 & 1 & 0\\
    \end{array}
  \right] x(t - 1) +
  \left[
    \begin{array}{ccc}
      0 & 0 & 0\\
      0 & 0 & 0\\
      a & 0 & 0\\
    \end{array}
  \right] x(t - 2) + v(t)
\end{equation*}

Then

\begin{align*}
  x_2(t) &= v_2(t) - ax_1(t - 1)\\
  x_3(t) &= v_3(t) + x_2(t - 1) + ax_1(t - 2)\\
  \Rightarrow x_3(t) &= v_2(t - 1) + v_3(t),
\end{align*}

and again $1 \npwgc 3$.
\end{example}

@@

\begin{definition}[Strongly Causal]
  \label{def:strongly_causal}
  We will say that a Granger-causality graph $\gcg$ is
  \textit{strongly causal} if there is at most one directed path between
  any two nodes.  Strongly Causal Graphs will be referred to as SCGs.
\end{definition}

@@

\begin{proposition}
  \label{prop:pwgc_anc}
  If $\gcg$ is a strongly causal DAG then $j \in \anc{i} \Rightarrow j \pwgc i$.
\end{proposition}
